---
title: "eda"
format: html
---

Statement of the research question and analysis objective:

What type of linear relationship exists between adjusted gross income and number of FEMA declared disasters in the years 2012 - 2022? The analysis objective is analysis.

Description of the data:

2 Most Important Variables - weighted_agi_stub, count

+------+--------------------+-------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+-----------------------------------------------------------+
| Year | COUNTYFIPS         | COUNTYNAME  | weighted_agi_stub                                                                                                                                                                        | designatedArea | count                                                     |
+======+====================+=============+==========================================================================================================================================================================================+================+===========================================================+
| Year | County FIPS number | County Name | SOI Tax Stats provide the distribution of tax returns on a scale of one to eight, each correlating to a tax bracket. This variable is the weighted average of tax bracket of the county. | County Name    | Number of FEMA declared disasters in the county that year |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 1 = Under \$1                                                                                                                                                                            |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 2 = \$1 under \$10,000                                                                                                                                                                   |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 3 = \$10,000 under \$25,000                                                                                                                                                              |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 4 = \$25,000 under \$50,000                                                                                                                                                              |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 5 = \$50,000 under \$75,000                                                                                                                                                              |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 6 = \$75,000 under \$100,000                                                                                                                                                             |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 7 = \$100,000 under \$200,000                                                                                                                                                            |                |                                                           |
|      |                    |             |                                                                                                                                                                                          |                |                                                           |
|      |                    |             | 8 = \$200,000 or more                                                                                                                                                                    |                |                                                           |
+------+--------------------+-------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+----------------+-----------------------------------------------------------+

Exploratory data analysis:

Weighted_agi v.s. Year, per county

```{r}
# plot_weighted_agi.R

library(tidyverse)

# Load the dataset
data_path <- "~/Documents/stat155/Project 1/data/processdata/weighted_agi.csv"
income_summary <- read.csv(data_path)

# Create output directory for plots
plot_dir <- "~/Documents/stat155/Project 2/plots"
dir.create(plot_dir, showWarnings = FALSE)

# Get unique counties
counties <- unique(income_summary$COUNTYNAME)

# Loop over counties and generate a plot for each
for (county in counties) {
  county_data <- income_summary %>%
    filter(COUNTYNAME == county)

  p <- ggplot(county_data, aes(x = Year, y = weighted_agi_stub)) +
    geom_line(color = "steelblue", size = 1) +
    geom_point(color = "darkred", size = 2) +
    labs(
      title = paste("Weighted AGI Stub Over Time:", county),
      x = "Year",
      y = "Weighted AGI Stub"
    ) +
    theme_minimal()

  ggsave(
    filename = paste0(plot_dir, "/", gsub(" ", "_", county), "_plot.png"),
    plot = p,
    width = 7, height = 5
  )
}

# Faceted plot: all counties together

# Create the faceted plot
faceted_plot <- ggplot(income_summary, aes(x = Year, y = weighted_agi_stub)) +
  geom_line(color = "steelblue") +
  geom_point(color = "darkred", size = 1.5) +
  facet_wrap(~ COUNTYNAME, scales = "free_y") +
  labs(
    title = "Weighted AGI Stub Over Time by County",
    x = "Year",
    y = "Weighted AGI Stub"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Save the faceted plot
ggsave(
  filename = "~/Documents/stat155/Project 2/agiplots/weighted_agi_facet_plot.png",
  plot = faceted_plot,
  width = 15, height = 10
)

```

Number of disaster v.s. Year, per county

```{r}
library(tidyverse)

# Read the merged data
mergeddata <- read.csv("~/Documents/stat155/Project 1/data/mergeddata.csv")

# Make sure count is set to 0 if NA (count of disasters)
mergeddata <- mergeddata %>%
  mutate(count = ifelse(is.na(count), 0, count))

# List of counties
county_list <- unique(mergeddata$COUNTYNAME)

# Create a folder to save graphs
output_dir <- "~/Documents/stat155/Project 2/disasterplots"
dir.create(output_dir, showWarnings = FALSE)

# Loop over each county
for (county in county_list) {
  
  # Filter data for that county
  county_data <- mergeddata %>%
    filter(COUNTYNAME == county)
  
  # Make the plot
  p <- ggplot(county_data, aes(x = Year, y = count)) +
    geom_point(color = "darkred", size = 2) +
    labs(title = paste("Disasters Over Time -", county),
         x = "Year",
         y = "Number of Disasters") +
    theme_minimal() +
     theme(
    plot.background = element_rect(fill = "white", color = NA),   # <- makes background white
    panel.background = element_rect(fill = "white", color = NA),  # <- also make panel white
    panel.grid.major = element_line(color = "grey80"),             # optional: lighter grid lines
    panel.grid.minor = element_blank()                             # optional: remove minor grids
  )
  
  # Save the plot
  ggsave(
    filename = paste0(output_dir, "/", county, "_disasters.png"),
    plot = p,
    width = 8,
    height = 5
  )
}

```
